Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-01 ms.tex
TDE section

Figure \ref{fig:bh_growth} (top panel) shows that across all black hole masses, Compton heating alone cannot prevent thermal instabilities from developing in a steady-state accretion flow ($\dot{M}_{\rm C} > \dot{M}_{\rm TI}$).  Thus, if the stellar population is sufficiently old (or the star formation sufficiently intermittent) then nuclear gas could undergo cyclic oscillations of thermal instability and high accretion rates, followed by quiescent periods once gas has been consumed and/or AGN feedback becomes effective.  Such cycles are proposed to occur on large scales and long timescales in elliptical galaxies (\citealt{Ciotti+10}), an idea supported, e.g. , by GALEX observations showing that a small fraction of early-type galaxies in the local universe have undergone recent ($< 1-2$ Gyr old) star formation episodes (\citealt{Donas+07}).  If such cyclic behavior also occurs on small, parsec scales (e.g.) then the large scatter in the accretion rates and hence X-ray luminosities shown in Fig \ref{fig:Miller} could also be understood by the wide range of gas masses experienced over the course of a cyclic episode between instabilities and periodic nuclear outbursts (e.g.~\citealt{Ciotti+10}).  In support of this possibility, we find that core galaxies are more likely to possess higher accretion rates and potentially be thermally unstable.  Past studies have found that core galaxies show a wide range of nuclear radio and X-ray luminosities (suggesting cyclic accretion episodes), while cusp galaxies are confined below a threshold (e.g., \citealt{Bender+89, Pellegrini99, Capetti&Balmaverde05}), consistent with thermally-stable picture.

The cyclic model predicts that high $L_{X}$ nuclei should be more common in galaxies with younger stellar ages, a result which appears in tension with the finding by \citet{Pellegrini10} that galaxies with younger stellar populations are observed to host less X-ray luminous nuclei, at least for stellar ages $<3$ Gyr.  An alternative explanation for this observation within a thermally stable accretion picture is that young stars provide higher stellar wind heating, reducing the mass accretion rate and hence nuclear X-ray luminosity.

\subsection{Local Growth of SMBHs }

The bottom panel of Figure~\ref{fig:bh_growth} shows the growth times for core and cusp galaxies, with parameters chosen for the average star formation history at each $\Mbh$ (see Figure~\ref{fig:vwSources} and~\ref{fig:eta}). Core and cusp galaxies are shown in green and blue respectively.  We also plot the maximum growth time for thermal
instability from equation~\eqref{eq:tgrow_min} in purple.

\begin{itemize}
\item{Fact that some low mass BHs appear to have nearly Eddington accretion rates, suggest an external origin - since the maximum Ia rate appears to be below this}
\item{Growth of low mass BHS, e.g. \citealt{Kauffmann&Heckman09}}
\end{itemize}

@@

Our results also have implications for jetted TDEs.  For black hole masses characterizing those responsible for {\it Swift} J1644+57, we predict accretion rates of $\dot{M} \sim$ corresponding to gas densities of $n \sim 0.1$ cm$^{-3}$ on radial scales.  This is similar to the value of $n \sim 1$ cm$^{-3}$ inferred for J1644+57 on radial scales $\sim 0.3$ pc (\citealt{Metzger+12}, \citealt{Berger+12}).

@@

\section{Death Row}

\subsection{X-ray Diagnostics of SMBH Occupation}
\label{sec:occ}
The free-free emission from steady state gas flows in low mass
galactic nuclei may serve as useful diagnostics of the presence or
absence of a SMBH.  We focus primarily on low-mass galaxies in this
subsection, both because this is the portion of parameter space in
which gas inflow will generally proceed in a steady-state fashion, and
because of recent observations suggesting that a nontrivial fraction
of these galaxies may lack nuclear SMBHs.  If this absence is
confirmed, it could have implications for galaxy evolution, SMBH
recoil distributions, and SMBH seed formation models.

We can approximate the gas temperature as

\begin{equation}
% T \approx \frac{\mu m_{\rm p} (\gamma-1)}{2\gamma k_{\rm B}}
% \left(\tilde{v}_{\rm w}^2 + GM_\bullet/r \right).
 \kb T \approx 2.1\times 10^{-25} \left(\tilde{v}_{\rm w}^2 + GM_\bullet/r
 \right) {\rm [cgs]}.
\end{equation}

In the absence of a central SMBH, we can expect to see an exponential
cutoff in the free-free X-ray spectrum at energies above

\begin{equation}
E_{\rm cut} = 0.32~{\rm keV} v_{500}^2,
\end{equation}

assuming $\gamma = 5/3$.  This cutoff will therefore be located in the
portion of the soft X-ray covered by {\it Chandra}'s ACIS
spectrometer.  By writing the free-free emissivity as

\begin{equation} 
\varepsilon_{\nu}^{\rm ff} = 6.8\times10^{-38}
\frac{\rm erg}{\rm s~cm^3~Hz} Z^2n_{\rm e}n_{\rm
i}T^{-1/2}\bar{g}_{\rm ff}\exp\left(\frac{-h\nu}{k_{\rm B}T}\right),
\end{equation}

where $Z$ is the atomic number of the ions, $n_{\rm e}$ and $n_{\rm
  i}$ are respectively the electron and ion number density, and
$\bar{g}_{\rm ff}$ is the Gaunt factor.  In the following we set
$n_{\rm e}=Y_{\rm e}n_{\rm i}$, and approximate
\begin{equation}
n_{\rm i} \approx \frac{\dot{M}}{(4\pi/3) \rs^2 \mu \mp}\sqrt{ \frac{r_{\rm s}}{ 2 G \Mbh} } \left( \frac{r}{\rs} \right)^{-n}.
\end{equation}

We now write the total spectral luminosity density seen in unresolved
X-ray observations above the cutoff energy as

\begin{align}
 L_{\nu}^{\rm ff} =& \int_0^{r_{\rm s}} 4\pi r^2 \varepsilon_{\nu}^{\rm ff}{\rm d}r \\
\approx & 2.8\times 10^{16} \, {\rm ergs \, s^{-1}
  \, Hz^{-1}} \left(\frac{\dot{M}}{10^{-5}M_\odot/{\rm
      year}} \right)^2 \left(\frac{\rs}{10^{-2}~{\rm pc}} \right)^{-1}
   \notag \\
& \times \left(\frac{h \nu}{0.5~{\rm keV}}\right)^{-3/2} \bar{Z}^2 \bar{g}_{\rm ff} Y_{\rm e} \notag \\
\approx & 3.9\times 10^{14} {\rm ergs~s^{-1}~Hz^{-1}}
\eta^2 \left(\frac{\Mbh}{10^{6}~{\Msun}} \right)^{1.96} \notag\\&\times
\left(\frac{h \nu}{0.5~{\rm keV}}\right)^{-3/2} v_{500}^{-2.8} \bar{Z}^2 \bar{g}_{\rm ff} Y_{\rm e} 
\end{align}

The second line of this equation makes the approximation that
$GM_\bullet/r$ dominates $\tilde{v}_{\rm w}^2$ throughout the
integrand, which will be true for energies modestly higher than
$E_{\rm cut}$.  The lower (upper) endpoint for this integral are
arbitrary so long as it is somewhat less (greater) than the radius at
which the observed frequency $\nu$ sees its exponential free-free
cutoff. We have also set $n=1$.

In general, these ``cut-off'' X-ray luminosities are quite small, and
even {\it Chandra} would have little hope of detecting them in any but
the nearest dwarf galaxies (e.g. the Magellanic clouds).  However, if
source confusion is not present, this technique could test for the
presence of IMBHs in very nearby dwarf galaxies, or more speculatively
Galactic star clusters.

>>>>

Figure \ref{fig:bh_growth} (top panel) shows that across all black hole masses, Compton heating alone cannot prevent thermal instabilities from developing in a steady-state accretion flow ($\dot{M}_{\rm C} > \dot{M}_{\rm TI}$).  Thus, if the stellar population is sufficiently old (or the star formation sufficiently intermittent) then nuclear gas could undergo cyclic oscillations of thermal instability and high accretion rates, followed by quiescent periods once gas has been consumed and/or AGN feedback becomes effective.  Such cycles are proposed to occur on large scales and long timescales in elliptical galaxies (\citealt{Ciotti+10}), an idea supported, e.g. , by GALEX observations showing that a small fraction of early-type galaxies in the local universe have undergone recent ($< 1-2$ Gyr old) star formation episodes (\citealt{Donas+07}).  If such cyclic behavior also occurs on small, parsec scales (e.g.) then the large scatter in the accretion rates and hence X-ray luminosities shown in Fig \ref{fig:Miller} could also be understood by the wide range of gas masses experienced over the course of a cyclic episode between instabilities and periodic nuclear outbursts (e.g.~\citealt{Ciotti+10}).  In support of this possibility, we find that core galaxies are more likely to possess higher accretion rates and potentially be thermally unstable.  Past studies have found that core galaxies show a wide range of nuclear radio and X-ray luminosities (suggesting cyclic accretion episodes), while cusp galaxies are confined below a threshold (e.g., \citealt{Bender+89, Pellegrini99, Capetti&Balmaverde05}), consistent with thermally-stable, steady-state picture.

The cyclic model predicts that high $L_{X}$ nuclei should be more common in galaxies with younger stellar ages, a result which appears in tension with the finding by \citet{Pellegrini10} that galaxies with younger stellar populations are observed to host less X-ray luminous nuclei, at least for stellar ages $<3$ Gyr.  An alternative explanation for this observation within a thermally stable, steady-state accretion picture is that young stars provide higher stellar wind heating, reducing the mass accretion rate and hence nuclear X-ray luminosity.

\subsection{Local Growth Time of SMBHs }

Low-mass SMBHs in the local universe ($M_{\bullet} \lesssim 3\times 10^{7}M_{\odot}$) are observed to be growing on a timescale comparable to the age of the universe, while the most massive SMBHs with $M_{\bullet} \gtrsim 10^{9}M_{\odot}$ possess local growth times which are more than 2 orders of magnitude longer (\citealt{Heckman+04}; \citealt{Kauffmann&Heckman09}).

The bottom panel of Figure~\ref{fig:bh_growth} shows the SMBH growth time, $t_{\rm grow} = M_{\bullet}/\dot{M}$, for each accretion rate shown in the top panel.  For black holes accreting steadily given the stellar wind heating associated with the average star formation history of their galaxies, we see that $t_{\rm grow}$ exceeds the Hubble time by 1-3 orders of magnitude across all black hole masses.  Steady accretion therefore cannot explain the  growth of low mass black holes, a fact which is not surprising given that approximately half of this growth occurs in AGN radiating within 10 per cent of their Eddington rate (\citealt{Heckman+04}).  Such high accretion rates likely instead require an source of gas external to the nuclear region, triggered either by galaxy mergers associated with the hierachical growth of structure or thermal instabilities on larger, galactic scales (e.g.~\citealt{Ciotti+10}).  That the growth time associated with thermally-unstable accretion (yellow line) exceeds the Hubble time across all SMBH masses highlights the fact that signifcant black hole growth in the local universe cannot result from thermally-stable steady accretion.

Gas blow-out by SN Ia cannot prevent the growth of low-mass black holes, as indicated by the low growth times $\ll t_{\rm h}$ allowed by Ia heating (green lines), although Ia heating could play in principle a role in capping the growth rate of $\gtrsim 10^{8}M_{\odot}$ black holes, again depending on the efficiency of Ia heating.

@@

Our results also have implications for the environments encountered by relativistic jets from TDEs.  For low-mass SMBHs  $M_{\bullet} \lesssim 10^{7}M_{\odot}$, characterizing those responsible for {\it Swift} J1644+57 (\citealt{Bloom:2011aa}), we predict gas densities of a $n \sim 0.3-3$ cm$^{-3}$ on radial scales of 0.3 pc for wind heating rates within the physically expected range $v_{\rm w} \sim 500-1000$ km s$^{-1}$ (Fig.~\ref{fig:profiles}).  These values are within the range estimated by modeling the radioa afterglow of {\it Swift} J1644+57 (\citealt{Metzger+12}, \citealt{Berger+12}).  In fact, the flattening of the density profile inferred by \citet{Berger+12} on radii scales $r \gtrsim 0.4$ pc (their Fig.~6) looks similar to that seen in our calculations of the gas density profile for core galaxies (Fig.~\ref{fig:profiles}, solid line).  The CNM density of the host of {\it Swift} J1644+57 was a factor of 30 times lower than that inferred by \citet{Quataert:2004a} for SgrA*, a fact we attribute to...{\bf BDM: whats going on?}  The implications of our models for CNM profiles for the diversity of radio emission from TDE jets will be explored further in future work.

@@
